<details open><summary class='drop'>Day 1</summary>
<br>
<details><summary class='ques2'>Question 1: How do you find the value of the 3rd row of domSpp?</summary>
```{r orgD1Ans, echo=F, eval=F, include=F}
#--------------------
#    Day 1 Answers
#--------------------
```

```{r, c105, echo=T, results='show'}
# If starting from scratch
library(tidyverse)
df <- data.frame(plot = c(1, 2, 3, 4, 5, 6, 'plot7'), #column for plot number
                 numtrees = c(24, 18, 17, 14, 31, 27, 15), 
                 numLive = c(21, 18, 16, 10, 29, 26, 14),
                 domSpp = c('red_maple', 'red_spruce', 
                            'red_spruce','white_cedar','white_pine',
                            'white_spruce', NA),
                 invasive_pres = c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE))


df[3, c("domSpp")] # option 1
df[3, 4] # option 2
```
</details>
<br>
<details><summary class='ques2'>Question 2: How would you make numtrees an integer?</summary> 
```{r, c106,echo=T, results='show'}
df$numtrees <- as.integer(df$numtrees)
str(df)
```
</details>
<br>

<details><summary class='ques2'>Question 3: How would you change the second row in invasive_pres from FALSE to TRUE?</summary> 
```{r, c106b,echo=T, results='show'}
df$invasive_pres[2] <- TRUE
head(df)
```
</details>
<br>

<details><summary class='ques2'>Question 4: How would you calculate the percent of trees that are dead in the df data frame?</summary>
```{r, c107,echo=T, results='hide'}
# base R version (2 options)
df$numDead <- df$numtrees - df$numLive
df$pctDead <- df$numDead/df$numtrees*100

df$pctDead <- with(df, (numtrees - numLive)/numtrees*100)

# tidyverse version (2 options)
df <- df %>% mutate(numDead = numtrees - numLive,
                    pctDead = numDead/numtrees*100)

df <- df %>% mutate(pctDead = (numtrees - numLive)/numtrees*100)
```
</details>
<br>

<details><summary class='ques2'>Question 5: What is the smallest number of dead trees observed in the df dataset?</summary>
```{r c107b, echo=T, results='show'}
min(df$numDead) # uses numdead column calculated for previous question. 
summary(df$numDead)[1] #Min. is the first record that summary reports for numeric columns
```
</details>
<br>

<details><summary class='ques2'>Question 6: How would you select only Invasive species records?</summary>
```{r, c108, echo=T, results='hide'}
# If starting from scratch
wetdat<-read.csv("./data/example_data/ACAD_wetland_data_clean.csv")

wetinv <- wetdat[wetdat$Invasive == TRUE, ] # base R bracket-based solution
wetinv2 <- subset(wetdat, Invasive == TRUE) # another base R solution
wetinv_tidy <- wetdat %>% filter(Invasive == TRUE) # tidyverse solution
```
</details>
<br>

<details><summary class='ques2'>Question 7: Which sites have both Arethusa bulbosa and Carex exilis present?</summary>
```{r, c109,echo=T, results='show'}
# Base R solution
wetspp <- wetdat[wetdat$Latin_Name %in% c("Arethusa bulbosa", "Carex exilis"),]
sort(unique(wetspp$Site_Name))

# Tidyverse solution
wetspp_tidy <- wetdat %>% filter(Latin_Name %in% c("Arethusa bulbosa", "Carex exilis")) 
sort(unique(wetspp_tidy$Site_Name)) # RAM-05 and SEN-01 have both species present
```
</details>
<br>

<details><summary class='ques2'>Question 8: How would you sort the wetdat by Ave_Cov (high to low) and Latin_Name?</summary>
```{r, c109b, echo=T, results='show'}
wetdat_sort <- wetdat[order(-wetdat$Ave_Cov, wetdat$Latin_Name), ] # base R
head(wetdat_sort)
wetdat_sort_tidy <- wetdat %>% arrange(-Ave_Cov, Latin_Name) #tidyverse
head(wetdat_sort_tidy)
```
</details>
<br>
</details>
<hr>
<details open><summary class='drop'>Day 2</summary>
<br>
<details><summary class='ques2'>Question 1: There's a Latin name that needs to be fixed (hint look for non-letter symbols). How would you find that record?</summary>

```{r orgD2Ans, echo=F, eval=F, include=F}
#--------------------
#    Day 2 Answers
#--------------------
```

```{r, c110, echo=T, results='show'}
sort(unique(wetdat$Latin_Name))[1:10] # Low-tech way: use your eyes to spot the "Alnus incana++". I truncated to first 10.
```
Hopefully you noticed Alnus incana++ when you sorted unique latin names. The ++ are errors that we want to remove (see Question 2). If you wanted to identify any non-alphabetical symbol in a column, and if your dataset was too big to spot all of them, you can use regular expressions like the code chunk below. The pattern in the grep is specifying that you want to return any non-alphabetical symbol (via alpha) in Latin_Name. Note that "^" is interpreted as NOT in regular expressions. To use the whole record for your search, rather than stopping at the first word (since Latin Names are usually 2 words), you add +$. The + specifies that you can have multiple matches within 1 record, and the \$ tells you to do the search to the end of the record.

```{r, c110bb, echo=T, results='show'}

# The regular expression way to find non-alphabetical symbols (now you know why I didn't cover this).
wetdat$Latin_Name[grep("[^[:alpha:]]+$", wetdat$Latin_Name)] # looks for anything not A-Z or a-z in entire record 
  # ^ and +$ are telling R to start with first word and check every following word within a record.
  # This took longer than I'd like to admit to make work, by the way.

```
</details>
<br>

<details><summary class='ques2'>Question 2: Now that you've found it, fix it, and check that you fixed it.</summary>
```{r, c110b, echo=T, results='show'}
wetdat$Latin_Name[wetdat$Latin_Name == "Alnus incana++"] <- "Alnus incana" # easy approach
wetdat$Latin_Name <- gsub("[+]", "", wetdat$Latin_Name) # gsub approach
wetdat$Latin_Name[grep("[^[:alpha:]]+$", wetdat$Latin_Name)] # no values returned 
sort(unique(wetdat$Latin_Name))[1:10] # using your eyes approach

```
</details>
<br>

<details><summary class='ques2'>Question 3: How would you take only the plots on MDI (eg Unit_ID = "ACAD_MDI") in the loc table and join them to their corresponding visits in the events table?</summary>
```{r, c111,echo=T, results='hide'}
# In case you need to load the tables
loc <- read.csv("./data/example_data/tbl_Locations_example.csv")
event <- read.csv("./data/example_data/tbl_Events_example.csv")  
stand <- read.csv("./data/example_data/tbl_Stand_Data_example.csv") 

# Filter then join
loc_MDI <- loc %>% filter(Unit_ID == "ACAD_MDI")
locev_MDI <- left_join(loc_MDI, event, by = "Location_ID")

# All in one via pipes:
locev_MDI2 <- loc %>% filter(Unit_ID == "ACAD_MDI") %>% 
                      left_join(event, by = "Location_ID")

```
</details>
<br>

<details><summary class='ques2'>Question 4: How would you join the stand data to the locev table (the one we made that dropped QAQC events)?</summary>
```{r, c111b, echo=T}
locev <- merge(loc, event[event$Event_QAQC == FALSE, ], all.x = TRUE, all.y = TRUE)
intersect(names(locev), names(stand))
locev_stand <- merge(locev, stand, by = "Event_ID", all.x = TRUE, all.y = TRUE) #base R full join
locev_stand_tidy <- full_join(locev, stand, by = "Event_ID") #tidyverse full join
```

</details>
<br>

<details><summary class='ques2'>Question 5: How would you calculate the min and max number of species by year?</summary>
```{r, c111c, echo=T}
head(wetdat)
wetmin_max <- wetdat %>% mutate(present = ifelse(Ave_Cov >0 & !is.na(Ave_Cov), 1, 0)) %>% 
                         group_by(Site_Name, Year) %>% 
                         summarize(numspp = sum(present), 
                                   .groups = "drop") %>% 
                         group_by(Year) %>% 
                         summarize(minspp = min(numspp),
                                   maxspp = max(numspp), 
                                  .groups = "drop")
wetmin_max
```
Note that first I had to sum up the number of species within a site (hence group_by(Site_Name, Year)). After I calculate that, I grouped by Year only, then found the min and max number of species recorded within a year.
</details>
<br>

<details><summary class='ques2'>Question 6: What percent of RAM sites had an invasive species present in 2012 and 2017?</summary>
```{r, c111d, echo=T}
#head(wetdat)
wetinv1 <- wetdat %>% filter(Site_Type == "RAM") %>% 
                      group_by(Site_Name, Year) %>% 
                      summarize(num_invspp = sum(Invasive), 
                                plot_inv = ifelse(num_invspp > 0, 1, 0), 
                                .groups = "drop")
head(wetinv1)
# Note plot_inv is to set plot to 1 or 0 even if multiple invasives are found on the plot.
wetinv <- wetinv1 %>% group_by(Year) %>% 
                      summarize(tot_inv_plots = (sum(plot_inv)/n())*100,
                                .groups = "drop")

wetinv 
```
Note that logical fields (ie TRUE/FALSE) are interpreted as numbers in R when you use sum. If the value is TRUE, R interprets it as 1. If it's FALSE, R interprets it as 0. So sum(Invasive) is summing all of the TRUEs in the Invasive column. 

</details>
<br>
</details>
<hr>
<details open><summary class='drop'>Day 3</details>
<br>
<details><summary class='ques2'>Question 1: How would you create a dataset that has a column for every sample year (ie 2013-2019), a row for every site, and the number of northern long-eared bats found as the value? </summary>

```{r orgD3Ans, echo=F, eval=F, include=F}
#--------------------
#    Day 3 Answers
#--------------------
```


```{r, c112, echo=T, results = 'hide'}
# In case you're starting from scratch
library(tidyverse)
bat_long <- read.csv("./data/example_data/fake_bats.csv")
bat_long <- bat_long %>% mutate(sppcode = toupper(paste0(substr(word(Latin, 1), 1, 3),
                                                         substr(word(Latin, 2), 1, 3))))

MYOSEP_sum <- bat_long %>% filter(sppcode == "MYOSEP") %>% 
                           group_by(Site, Year) %>% 
                           summarize(num_bats = sum(ifelse(!is.na(Latin), 1, 0)), 
                                     .groups = 'drop') 

head(MYOSEP_sum) # Note only the site and years where MYOSEP was found are listed. Next lines add those back before pivot
bat_site_year <- bat_long %>% expand(Site, Year)

MYOSEP_comp <- bat_site_year %>% left_join(., MYOSEP_sum, by = c("Site", "Year")) 
MYOSEP_comp$num_bats[is.na(MYOSEP_comp$num_bats)] <- 0 # Replace all the NAs created by left join with 0

MYOSEP_wide <- MYOSEP_comp %>% arrange(Site, Year) %>% 
                               pivot_wider(names_from = Year, values_from = num_bats, values_fill = 0)
```
```{r, c112b, echo=T}
MYOSEP_wide
```

</details>
<br>

<details><summary class='ques2'>Question 2: How would you create a wide matrix with all protected species as columns, sites and years sampled as rows, and percent cover as the value?</summary>
```{r, c113,echo=T}
# If you're starting from scratch
wetdat <- read.csv("./data/example_data/ACAD_wetland_data_clean.csv")
wetdat_prep <- wetdat %>% mutate(sppcode = toupper(paste0(substr(word(Latin_Name, 1), 1, 3),
                                                          substr(word(Latin_Name, 2), 1, 3)))) 

wetdat_dist <- wetdat_prep %>% distinct(Site_Name, Year) # makes tibble of all site, year combinations, but doesn't add any like expand

wetdat_wide <- wetdat_prep %>% filter(Protected == TRUE) %>% 
                               select(Site_Name, Year, sppcode, Ave_Cov) %>% 
                               pivot_wider(names_from = sppcode, values_from = Ave_Cov, values_fill = 0) %>% 
                               left_join(wetdat_dist, ., by = c("Site_Name", "Year")) 
wetdat_wide[, c("AREBUL", "POGOPH", "CALTUB")][is.na(wetdat_wide[, c("AREBUL", "POGOPH", "CALTUB")])] <- 0

```
Note that I used distinct() instead of expand() because not every site is sampled every year. I used distinct to add site/year combinations that were sampled but where protected species weren't found. Expand would have added a record for every year in the data to every site. 

</details>
<br>
<details><summary class='ques2'>Question 3: How would you create a wide matrix for ACAD plots in 2019 that has plots for rows, and exotic species as columns?</summary>

```{r, c114,echo=T, results='hide'}
# Starting from scratch
library(forestNETNarch)
importCSV("./data/NETN_forest_csvs")
acad_exo <- makeSppList(park = "ACAD", from = 2019, to = 2019, speciesType = "exotic") 
acad_exo <- acad_exo %>% mutate(present = ifelse(Exotic == TRUE & !is.na(Exotic), 1, 0))
head(acad_exo)
```
```{r, c114b, echo=T}
# Now to answer the question
# Old school spread version
acad_wide_spread <- acad_exo %>% select(Plot_Name, Latin_Name, present) %>% 
                                 arrange(Latin_Name, Plot_Name) %>% 
                                 spread("Latin_Name", "present", fill = 0)

# Newer pivot version
acad_wide_pivot <- acad_exo %>% select(Plot_Name, Latin_Name, present) %>% 
                                arrange(Latin_Name, Plot_Name) %>% 
                                pivot_wider(names_from = "Latin_Name", 
                                            values_from = "present",
                                            values_fill = 0) 

head(acad_wide_pivot)
```
</details>
<br>

<details><summary class='ques2'>Question 4: Take the acad_final graph and make 2 more changes to it.</summary>

You may have chosen other things to do than I did. Here are a couple of changes to the ggplot bar chart
```{r, c115, echo=T, results='hide', eval=FALSE}
# Starting from scratch
library(forestNETN)
importCSV("./data/NETN_forest_csvs")
acad_regen <- joinRegenData(park = "ACAD", from = 2016, to = 2019, canopyForm = "canopy")

acad_regen2 <- acad_regen %>% group_by(Plot_Name) %>% 
                              summarize(seed15_30 = sum(seed15.30),
                                        seed30_100 = sum(seed30.100),
                                        seed100_150 = sum(seed100.150),
                                        seed150p = sum(seed150p),
                                        sapling = sum(sap.den),
                                        .groups = 'drop') %>% 
                              arrange(Plot_Name)

acad_regen_long <- acad_regen2 %>% pivot_longer(-Plot_Name, 
                                                names_to = "size_class", 
                                                values_to = "density")

# Now we calculate the mean and se for each size class at the park level
acad_stats <- acad_regen_long %>% group_by(size_class) %>% 
                                  summarize(mean_dens = mean(density),
                                            se_dens = sd(density)/sqrt(n()),
                                            .groups = 'drop')

acad_stats$size_class_fact <- ordered(acad_stats$size_class, 
                                      levels = c('seed15_30', 'seed30_100', 
                                                 'seed100_150', 'seed150p', 'sapling'))
```
```{r c115bb, echo=T}
# Now to answer the question
ggplot(data = acad_stats, aes(x = size_class_fact, y = mean_dens))+
              geom_bar(stat = 'identity', fill = "steelblue4")+
              geom_errorbar(aes(ymin = mean_dens - se_dens, ymax = mean_dens + se_dens),
                            width = 0.2, lwd = 0.7)+
              labs(x = "Regeneration Size Class", y = "Stems per sq.m")+
              theme_FHM()+
              theme(axis.text.x = element_text(angle = 45, hjust = 1))+ #hjust is centering horizontally
              scale_x_discrete(labels = c('15-30cm', '30-100cm', '100-150cm', '>150cm', '1-10cm DBH'))
             
```

Note that width = 0.2 in the geom_errorbar is the width of the caps on the bars. lwd is the weight of the line. 
</details>
<br>

<details><summary class='ques2'>Question 5: Filter the acad_ph_final data frame to only include 2019 data and create a plot with months as the X axis, and includes points, a loess smooth with span = 1, and a facet on site.</summary>
```{r, c115b, echo=T}
# Starting from scratch
library(NCRNWater)

netnwd<-importNCRNWater(Dir = "./data/NETN_water_data", 
                        Data = "Water Data.csv", 
                        MetaData = "VizMetaData.csv")

acad_all_sites <- data.frame(sitecode = getSiteInfo(netnwd, parkcode = "ACAD", info = "SiteCode"),
                             type = getSiteInfo(netnwd, parkcode = "ACAD", info = "type"),
                             fullname = getSiteInfo(netnwd, parkcode = "ACAD", info = "SiteName"))

acad_lakes <- acad_all_sites %>% filter(type == "Lake")  #list of site codes and full names for ACAD lakes

acad_ph <- getWData(netnwd, parkcode = "ACAD", years = 2006:2019, charname = 'pH')
acad_lakes_ph <- acad_ph %>% filter(Site %in% acad_lakes$sitecode)
annual_lakes <- data.frame(table(acad_lakes_ph$Site)) %>% filter(Freq > 90) 

acad_ph2 <- acad_lakes_ph %>% filter(Site %in% annual_lakes$Var1) %>% 
                              mutate(month = format(Date, "%m"),
                                     year = format(Date, "%Y"))

acad_ph_final <- left_join(acad_ph2, acad_lakes[, c("sitecode", "fullname")], 
                           by= c("Site" = "sitecode")) # how to join if columns have diff. names

#head(acad_ph_final)

# Now to answer the question
acad_ph_2019 <- acad_ph_final %>% filter(year == 2019) %>% mutate(mon_abr = format(Date, "%b"))

plot19 <- ggplot(acad_ph_2019, aes(x = Date, y = Value))+
             geom_point()+
             labs(y = "pH")+
             geom_smooth(method = 'loess', span = 1, se = FALSE)+
             scale_x_date(date_breaks = "1 month",
                          date_labels = "%b")+ #%b is months names abbreviated
             facet_wrap(~fullname, ncol = 4)+
             theme_bw()

plot19
```

</details>
<hr>