<h2>Day 1</h2>
<details><summary class='ques2'>Question 1: How do you find the value of the 3rd row of domSpp?</summary>
```{r, c105,echo=T, results='show'}
df[3, c("domSpp")]
```
</details>
</br>
<details><summary class='ques2'>Question 2: How would you make numtrees an integer?</summary> 
```{r, c106,echo=T, results='show'}
df$numtrees <- as.integer(df$numtrees)
str(df)
```
</details>
</br>
<details><summary class='ques2'>Question 3: How would you calculate the percent of trees that are dead?</summary>
```{r, c107,echo=T, results='hide'}
# base R version (2 options)
df$numdead <- df$numtrees - df$numLive
df$pctdead <- df$numdead/df$numtrees*100

df$pctdead <- with(df, (numtrees - numLive)/numtrees*100)

# tidyverse version (2 options)
df <- df %>% mutate(numdead = numtrees - numLive,
                    pctdead = numdead/numtrees*100)

df <- df %>% mutate(pctdead = (numtrees - numLive)/numtrees*100)
```
</details>
</br>
<h2>Day 2</h2>
<details><summary class='ques2'>Question 4: How would you select only the plots in ACAD and MIMA from the loc table?</summary>
```{r, c108, echo=T, results='hide'}
loc_ACAD <- loc[loc$Unit_ID %in% c("ACAD", "MIMA") & !is.na(loc$Unit_ID),]

loc_VAFO_HOFU_tidy <- loc %>% filter(Unit_ID %in% c("ACAD", "MIMA"))

```
</details>
</br>
<details><summary class='ques2'>Question 5: How many plots are in MABI and SARA (based on loc table)?</summary>
```{r, c109,echo=T, results='show'}
loc_MABI_SARA_tidy <- loc %>% filter(Unit_ID %in% c("MABI", "SARA")) %>% select(Unit_ID, Plot_Number) %>% unique()
nrow(loc_MABI_SARA_tidy)

loc_MABI_SARA_tidy_no_nas <- loc %>% filter(Unit_ID %in% c("MABI", "SARA") & !is.na(Plot_Number)) %>% select(Unit_ID, Plot_Number) %>% unique()
nrow(loc_MABI_SARA_tidy_no_nas)

```

</details>
</br>
<details><summary class='ques2'>Question 6: How would you calculate the min and max for seed_den_m2?</summary>
```{r, c110,echo=T, results='show'}
reg_by_park2 <- regdf %>% group_by(Unit_Code) %>% 
                          summarize(min_seed_dens = min(seed_den_m2),
                          max_seed_dens = max(seed_den_m2))

reg_by_park2
```
</details>
</br>
<details><summary class='ques2'>Question 7: How would you take only QA/QC events and join the loc table to QA/QC events only?</summary>
```{r, c111,echo=T, results='hide'}
eventQAQC <- event %>% filter(Event_QAQC == TRUE)
locevent_QAQC <- merge(loc, eventQAQC, by="Location_ID", all.x = F, all.y = T)
```
This is technically a right join, because I'm telling R to take all of the records from the right table (eventQAQC) and only join records in left (loc) table that match. This is just a style preference. I prefer to keep higher level tables to the left of lower level tables (eg tables with 1 to many relationships). The loc (tbl_Locations) table is higher level than the event (tbl_Events) table, because the loc table has the 1 plot record and the event table has the many visits for the plot records. If this is confusing, don't worry about it. You'll just notice that it's the order I generally organize data in my code.  
</details>
</br>
<h2>Day 3</h2>
<details><summary class='ques2'>Question 8: How many plots were sampled in ACAD in 2018?</summary>
```{r, c112,echo=T}
acad <- joinLocEvent(park = "ACAD", from = 2018, to = 2018)
length(unique(acad$Plot_Number))
```
</details>
</br>

<details><summary class='ques2'>Question 9: Which plot in ACAD has the highest density of live trees in the 90-99.9cm size class for any year?</summary>
```{r, c113,echo=T}
acad_tree_dist <- sumTreeDBHDist(park = "ACAD", status = 'live')

acad_tree_max70 <- acad_tree_dist %>% arrange(desc(d70_79.9))
head(acad_tree_max70) # ACAD-124

# If you wanted to pull out the top record for that size class, you can also do one of the two options below
acad_top_70 <- acad_tree_dist[which.max(acad_tree_dist$d70_79.9),] #The ugly base R approach 
acad_top_70_tidy <- acad_tree_dist %>% arrange(desc(d70_79.9)) %>% slice(1) #The tidyverse approach
head(acad_top_70)
```
</details>
</br>
<details><summary class='ques2'>Question 10: How would you create a wide matrix for ACAD plots in 2019 that has plots for rows, and exotic species as columns?</summary>

```{r, c114,echo=T}
acad_exo <- makeSppList(park = "ACAD", from = 2019, to = 2019, speciesType = "exotic") 
acad_exo <- acad_exo %>% mutate(present = ifelse(Exotic == TRUE & !is.na(Exotic), 1, 0))
head(acad_exo)

# Old school spread version
acad_wide_spread <- acad_exo %>% select(Plot_Name, Latin_Name, present) %>% 
                                 arrange(Latin_Name, Plot_Name) %>% 
                                 spread("Latin_Name", "present", fill = 0)

# Newer pivot version
acad_wide_pivot <- acad_exo %>% select(Plot_Name, Latin_Name, present) %>% 
                                arrange(Latin_Name, Plot_Name) %>% 
                                pivot_wider(names_from = "Latin_Name", 
                                            values_from = "present",
                                            values_fill = 0) 

head(acad_wide_pivot)
```
</details>
</br>
<h2>Day 4</h2>
<details><summary class='ques2'>Question 11: Take the gett_final graph and make 2 more changes to it.</summary>

You may have chosen other things to do than I did. Here are a couple of changes to the ggplot bar chart
```{r, c115,echo=T}

 ggplot(data = acad_stats, aes(x = size_class_fact, y = mean_dens))+
              geom_bar(stat = 'identity', fill = "steelblue4")+
              geom_errorbar(aes(ymin = mean_dens - se_dens, ymax = mean_dens + se_dens), width = 0.2, lwd = 0.7)+
              labs(x = "Regeneration Size Class", y = "Stems per sq.m")+
              theme_FHM()+
              theme(axis.text.x = element_text(angle = 45, hjust = 1))+ #hjust is centering horizontally
              scale_x_discrete(labels = c('15-30cm', '30-100cm', '100-150cm', '>150cm', '1-10cm DBH'))
             
```

Note that width = 0.2 in the geom_errorbar is the width of the caps on the bars. lwd is the weight of the line. 
</details>
</br>