#### Project and File Setup{.tabset}
<h2>Project and File Setup</h2>
First thing we're going to do is start an R project for this training and set up the same file structure, which we'll continue to use the rest of the training. To set up a project, open R Studio and go to: </br> </br><div style="text-indent:20px"> <h3>File > New Project > New Directory > New Project</h3></div></br> 
Find the directory you're going to work from (other than avoiding saving to your desktop, anywhere is probably fine), and name the directory the R_training. 

<p class='code'>Now run the code chunk below to set up the same file structure.</p> 

```{r, c3, echo=TRUE}
# Step 1m 
subfolders<-c("figures", "tables", "rscripts", "shapefiles")

invisible(lapply(subfolders, function(x){
  if(!dir.exists(x)){dir.create(paste(getwd(), x, sep = '/'))}
  }))

if(!dir.exists("./data")){dir.create('./data')} # creating a data folder to store NETN data tables if it doesn't already exist. 
```

Now we want to download the NETN data we're going to explore later in the week. Unfortunately there's a firewall between Sharepoint/Teams and R, so we can't download the data directly in R. To download the NETN data, go to The ACAD_R_Training_Info team and download <a href="https://doimspp.sharepoint.com/sites/ACAD_R_Training_Info/Shared%20Documents/General/data/data.zip">data.zip</a> in the files channel. Copy the data.zip file to the data folder you created in the previous step. Then run the code below. 

```{r, c3data, echo=TRUE}
list.files("./data") # check that you have data.zip in your data folder
unzip("./data/data.zip", exdir = "./data")
list.files("./data") # check that the unzip worked. There should be multiple folders with datasets within each.
```

<ul><li>This approach makes sharing code easier, because the parent directly can vary across machines, as long as the project has the same subdirectory. </li>
<li>The last thing to mention here is the # in the code chunk above. The # comments out anything after it on the same line, so R knows not to try to run it. Commenting your code allows you to document what you're doing, so it's easier for someone else (or you 3 months later) to follow your code. </li></ul>
