#### Data Summary I{.tabset}

<h2>Data Summary I</h2>
Using the df we created in the previous section, let's perform some basic summary statistics. Base R has a number of summary statistics built into it that are really useful, including mean, median, min, max, and sd (for standard deviation).  

```{r orgD1M5, echo=F, eval=F, include=F}
#--------------------
#    Day 1 Mod 5
#--------------------
```

```{r, c15, echo=T, results='show'}
min(df$numtrees)
max(df$numtrees)
mean(df$numtrees)
range(df$numtrees) #gives min and max
```

The summary() function in R is another really useful function that can take just about any named Data object in your global environment and provide you a meaningful summary. For data frames, the summary() function gives different/meaningful summaries for each column depending on type (e.g. numeric vs. character). The summary() function also works well for summarizing model output, like from ANOVA, linear regression, etc. 
```{r, c16, echo=T, results='show'}
summary(df)
```

You can also calculate new columns from existing columns in your data.
```{r, c17, echo=T, results='show'}
df$numDead <- df$numtrees - df$numLive # base R version
df <- df %>% mutate(numDead = numtrees - numLive) # tidyverse version of the same operation.
# note if the 2nd line doesn't run, load the tidyverse library (library(tidyverse)) first.
```

The <b>mutate</b> function is in the <i>dplyr</i> package of the tidyverse. The mutate function is used to create a new column in an existing data frame, and always returns the same number of records as the existing data frame. In this example, mutate creates a new column called numdDead, and tells it to equal numtrees - numLive. The results are identical to the base R version in the line above. Mutate just spares you having to write df$ over and over. 

<p class='ques'>Question 4: How would you calculate the percent of trees that are dead in the df data frame? </p>

<p class='ques'>Question 5: What is the smallest number of dead trees observed in the df dataset? </p>

<hr>