#### Getting Help{.tabset} 
<h2>Getting Help</h2>
There are a number of options to get help with R. If you're trying to figure out how to use a function, you can type ?function_name. For example <i>?plot</i> will show the R documentation for that function in the Help panel. <i>?dplyr::filter</i> works too, if you need to connect the package to the function (eg filter is also a function in stats, which is loaded as part of base R). 

```{r, c4, echo=T, eval=F}
?plot
?dplyr::filter
```

To view all of the functions and links to their individual help files, type in help(package=packagename). For example, the code below will bring up the help files for the dplyr package.
```{r, c5, echo=T, eval=F}
help(package=dplyr)
```

Online resources include <a href="https://stackexchange.com/">Stackexchange</a>, and <a href="https://stackoverflow.com//">Stackoverflow</a>. For tidyverse-related packages/questions, there's really good documentation for each package and function <a href="https://tidyverse.tidyverse.org/"> here</a>. Google searches are usually my first step, and I include "in R" and the package name if applicable in every search related to R code. 

<hr>