#### Joining data frames

<details open><summary class='drop'>Basics of joining tables</summary>
In NETN, most of our data are stored in relational databases, where plot-level data that never change (eg X, Y coords) are stored in 1 table (tbl_Location) and data related to each visit (also called events) are stored in other tables, we often have to join the tables together to work with them in R. To do that, we use a column that is found in each table called a primary key, that allows you to match the correct records in one table to the other. In our database, we typically use GUIDs (globally unique identifiers) as our keys, which I'll point out in the next section. 

There are different flavors of joining tables. The main joins we use are: 
<ul>
  <li><b>Left joins:</b> take all of the records in the primary key on the left side data frame (ie the first data frame specified in 
  the function) and find one or more matching records in the right side data frame (ie the second data frame specified). Any records in 
  the left data frame that don't have a match in the right data frame will still be included in the final result, but all of the columns 
  coming from the right data frame will be NA (blank). Records on the right data frame without a match in the left data frame are 
  dropped from the result.</li>
  <li><b>Right joins:</b> basically the same as left joins, except the second dataset includes all records and the first only includes 
  records with matching primary keys.</li> 
  <li><b>Semi joins:</b> only take records from both left and right datasets that have matching primary keys.</li>
  <li><b>Full joins:</b> take all of the records from both sides of the join. Records on both sides without matches are included in 
  the results with blanks in the cells that came from the table missing a match.</li>
</ul>

In practice, I use left and full joins the most often, so I'll focus mostly on those in the rest of this module. We'll try this out on simple datasets first, then we'll build up to more complex examples with our forest data. Another great example with visuals can be found <a href="https://mgimond.github.io/ES218/Week03c.html">here</a>.
```{r, c34a, echo=T, results ='hide'}
tree <- data.frame(plot = c("p1","p1","p1","p2","p2","p3","p4","p4"), 
                  treenum = c(1, 2, 3, 11, 12, 21, 32, 33),
                  species = c("ACERUB" ,"PINSTR", "PINSTR", "QUERUB", "PICRUB", 
                              "PICRUB", "PICRUB", "ACERUB"))
tree

tree_data <- data.frame(plot = c("p1", "p1", "p1", "p1", "p1","p1", "p2", "p2", "p5", "p7"),
                        year = c(2012, 2012, 2012, 2016, 2016, 2016, 2012, 2012, 2016, 2016),
                        treenum = c(1, 2, 3, 1, 2, 3, 11, 12, 51, 71),
                        DBH = c(12.5, 14.5, 16.1, 12.8, 14.9, 17.2, 28.1, 35.4, 36.1, 45.2))

tree_data

tree_full_join <- full_join(tree, tree_data, by = c("plot", "treenum"))
tree_left_join <- left_join(tree, tree_data, by = c("plot", "treenum"))
```
```{r, c34b, echo=T}
nrow(tree_full_join)
nrow(tree_left_join)
print(tree_full_join)
print(tree_left_join)
```
Note the left join has fewer records, and doesn't include p5 and p7 in the plot column from tree_data. There's also no tree data for plots p3 and p4 because there wasn't any data in the tree_data that matched. The full join includes plots and tree numbers from both tables, but there are NAs where there wasn't a matching record in one of the tables.

</details>
<br>
<details open><summary class='drop'>Advanced table joins</summary>
Let's load some example forest data, and use the primary keys to join them together.

```{r, c32, echo=T, results='hide'}
library(tidyverse)
loc <- read.csv("./data/example_data/tbl_Locations_example.csv") # plot level data for forest plots in NETN
event <- read.csv("./data/example_data/tbl_Events_example.csv") # data on each visit in NETN forest plots
stand <- read.csv("./data/example_data/tbl_Stand_Data_example.csv") # stand data collected in plots
```
```{r, c33, echo=T, eval=F}
View(loc) #Location_ID
View(event) #Location_ID links to tbl_Location; Event_ID links other visit-level tables to this table
View(stand) #Event_ID links stand table to event table. 
```
```{r, c34, echo=T}
intersect(names(loc), names(event)) # way to check columns two tables have in common.

```
The primary key to join the loc and event table is Location_ID, which the intersect function also shows is a column they have in common. Notice that while the loc table has a lot of important information in it, there's no information about when and how often a plot has been sampled. The event table, on the other hand has all of the dates a plot was sampled, but it doesn't have the park or plot columns to determine which parks/plots those dates are associated with. We need to join the data from both tables (I'm using data frame and table interchangeably here) together, to determine that.

```{r, c35, echo=T, results='hide'}
# base R version
names(loc) # to help me pick the columns to include in the join
names(event)
loc_event_full <- merge(loc, event, by = "Location_ID", all.x = TRUE, all.y = TRUE) # base R full join
loc_event_full_tidy <- full_join(loc, event, by = "Location_ID") # tidyverse full join

```
```{r, c36, echo=T, eval=F}
View(loc_event_full) # Now you can see all the times a plot has been sampled. You'll also see that some plots 
# don't have sample dates. That's because the loc table includes plots that were rejected and never sampled.
```
```{r, c37, echo=T}
loc_event_left <- merge(loc, event, by = "Location_ID", all.x = TRUE, all.y = FALSE)
loc_event_left_tidy <- left_join(loc, event, by = "Location_ID")
nrow(loc_event_full) # same number of rows because all events have a location record
nrow(loc_event_left) # same number of rows because all events have a location record

# tidyverse version
loc_event_full_tidy <- full_join(loc, event, by = "Location_ID")
nrow(loc_event_full_tidy) # same number of rows because all events have a location record

loc_event_left_tidy <- left_join(loc, event, by = "Location_ID")
nrow(loc_event_left_tidy) # same number of rows because all events have a location record

```
Note that the merge() function is the base R way of joining databases. The all.x = TRUE and all.y = TRUE makes it a full join. Left join is all.x = TRUE, all.y = FALSE, and right join is all.x = FALSE, all.y = TRUE. I tend to use merge over the tidyverse join approach, because it's a bit faster when you have huge datasets, and it's what I've gotten used to. The exception, which I'll show below, is if I'm chaining multiple steps that include a join via pipes.

Let's filter the location data to only include ACAD plots, and the event table to only include non-QAQC plots (we resample 5% of plots per year; QAQC=TRUE represents the 2nd visit). 

```{r, c38, echo=T, results='hide'}
# Base R approach
locev <- merge(loc[loc$Park_Code == "ACAD", ], 
               event[event$Event_QAQC == FALSE, ], 
               all.x = TRUE, all.y = FALSE)

# Tidyverse approach
loc_ACAD <- loc %>% filter(Park_Code == "ACAD")
locev_tidy <- event %>% filter(Event_QAQC == FALSE) %>% 
                        left_join(loc_ACAD, ., by = "Location_ID")

nrow(locev) 
nrow(locev_tidy)
```
```{r, c39, echo=T, eval=F}
View(locev)
View(locev_tidy)
```
Note the use of "." above. Functions in the tidyverse all assume the dataset to the left of the pipe will go into the first argument in the function following the pipe. When that's not the case, such as in the left_join above, you can use the "." to specify where to use the dataset. 

<p class="ques"> Question 3: How would you take only the plots on MDI (eg Unit_ID = "ACAD_MDI") in the loc table and join them to their corresponding visits in the events table?</p>

<p class="ques"> Question 4: How would you join the stand data to the locev table?</p>
</details>
<br>
<hr>
