#### Types of Data{.tabset}
<details open><summary class='drop'>Getting Started</summary>
Now we're going to start coding. The best way to work through these lessons will be to take the code in the R_Training_code.html doc and copy it in to an R Script in your R_training.Rprj. Go ahead and go to File > New File > R Script (the white rectangle with a green Plus just below the word File will do the same.), and name the script Day_1.R, and save it to your rscripts subfolder. As you get  more comfortable coding in R, try pushing yourself to type the commands, rather than copying and pasting. You'll learn to code faster by typing, even if you're just mimicking what you see in the training document. 

Before we go too far, let's make sure you have the tidyverse package loaded by running the code chunk below. This is the only dependency we'll have for today's code.
```{r, c5b, echo=T, eval=F}
library(tidyverse)
```
</details>
</br>
<details open><summary class='drop'>Types of Data</summary>
<p>There are multiple types of data in R, including vectors, lists, arrays, matrices, dataframes and tibbles. We will spend most of our time with dataframes, and won't worry about the other data types. Tibbles are the tidyverse version of a dataframe, but they're essentially the same thing as a dataframe with a few added bells and whistles. 

Dataframes are typically organized with rows being records and columns being variables (also called tidy data). The column variables can be numbers, text, or logical (TRUE/FALSE). Before we go any further, let's create a simple dataframe to look at and perform some basic functions. </p>

<p class="code">Copy the code below into the console and run it.</p>
```{r, c7, echo=T}

df <- data.frame(plot = c(1, 2, 3, 4, 5, 'plot7'), #column for plot number
                 numtrees = c(24, 18, 17, 14, 31, 27), 
                 numLive = c(21, 18, 16, 10, 29, 26),
                 domSpp = c('red_maple', 'willow_oak', 'loblolly_pine','sweet_gum','tulip_poplar', NA),
                 invasive_pres = c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE))

print(df)
```

Here we're creating a dataframe object named <b>df</b> and using <b> <- </b> to assign the stuff on the right to <b>df</b>. The df dataframe has 5 columns and 6 rows. Similar to Excel, R treats numbers and text differently, and some functions will only work with numbers and some will only work with text. 
A couple of points about working with text columns in R dataframes:
<ul>
<li>Text is called a string in R. If you have any text in a column, R automatically makes it a string. </li>
<li>There are 2 types of strings in R: character (abbreviated chr) and factor. </li>
<li>Older versions of R (i.e., R < 4.0) defaulted strings to factors. The newest R version defaults to characters, which saves us a lot of headaches (more on this later). </li>
<li>When you're dealing with strings, you usually have to put quotes around them. Note that 99% of the time, it doesn't matter whether you use single or double quotes, as long as the thing you're quoting is consistent.</li>
<li>The pound sign (#) is used to tell R not to run the text following the #. This is also called commenting out text, and is a way to leave comments that won't break your code.</li></ul>

Columns that are entirely numbers (blanks are okay too) are numeric. Integers are numeric whole numbers that will only return integers. That's usually not a problem, but if it ever becomes a problem, the as.numeric() function fixes it. Logical is the last data type, and only consists of TRUE and FALSE. 

Finally NA is a blank or missing value in your data. Note that the last row in domSpp is NA (stands for not available).  
</details>
</br>
<details open><summary class='drop'>Selecting rows and columns in R</summary>
There are 2 main approaches in base R for selecting/working with rows and columns- the $ and [ , ]. 
<ul>
  <li> The \$ is used to specify a dataframe name and one of its fields. For example df$plot is referring to the plot column in the df
  dataframe.</li>
  <li> The brackets [ , ] allow you to select columns and rows. The left side of the comma in brackets handles rows, and the right side of the bracket
  deals with columns, like: df[row numbers, column numbers]. </li>
</ul> 

<p class='code'>To better understand this, run the code below:</p>
```{r, c8, echo=T, eval=F}
df$plot # List the values within the column plot for the df dataframe
df[ , c('plot')] # Bracket version of the same thing
df[ , 1] # Bracket version of showing the first column, which is plot
df[2, 1] # Selecting 2nd row of first column

```

<p class='ques'>Question 1: How do you find the value of the 3rd row of domSpp? Answer is in the Answers tab.</p>
</details>
</br>
<details open><summary class='drop'>Check and fix data structure</summary>
<p class='code'>To check the structure of your data, run the code chunk below.</p>
```{r, c9, echo=T}
str(df) # to view how R is defining each column
head(df) # to view the first 6 rows of df
names(df) # to view the column names and their order
```

Note the different data types of the columns. The plot column is called a chr, and not a number because of the last value. There's also a blank in domSpp. Let's modify the dataset so that plot reads in as numeric, and replace the missing value in domSpp with "white oak". 

<p class='code'>Copy (or type) each section of code below and run in your console.</p>
```{r, c10, echo=T}
# The long way (note that I renamed the object here)
df2 <- data.frame(plot = c(1, 2, 3, 4, 5, 7), 
                 numtrees = c(24, 18, 17, 14, 31, 27), 
                 numLive = c(21, 18, 16, 10, 29, 26),
                 domSpp = c('red_maple', 'willow_oak', 'loblolly_pine','sweet_gum','tulip_poplar', 'white_oak'),
                 invasive_pres = c(TRUE, FALSE, FALSE, FALSE, FALSE, TRUE)) 

# A shorter way
df$plot[6] <- 7 #takes the 6th row of the plot column and changes it to 6
df$domSpp[6] <- "white_oak" #takes the 6th row of domSpp and replaces it with "white_oak"

# Another shorter way
df[6, c("plot")] <- 7 #takes the 6th row of the plot column and makes it 6
df[6, c('domSpp')] <- "white_oak" #takes the 6th row of the domSpp column and makes it "white_oak"

# Another shorter way
df[6, 1] <- 7 #takes the 6th row of the first column and makes it 6
df[6, 4] <- "white_oak" #takes the 6th row of the 4th column and makes it "white_oak"

# Even shorter-way
df[6, c(1, 4)]<-c(7, "white_oak") #Takes the 6th row and the first and 4th column, which happen to by plot and domSpp.
# Then the first column gets 7, and the 4th gets "white oak"
```

This is merely to show that there are usually 5 different ways to do something in R, and the best way is the one you can get to work. At least that's my philosophy for about 75% of the coding I do. The other 20% is about having code I can pick up a year later and still understand what I did. The last 5% is about coding for performance (e.g. if it's a big dataset).

I should mention that when you use numbers instead of column names, your code is pretty fragile to changes in column order (eg, you drop a column). Row numbers are similarly fragile, if you reorder your data. It's better to use names where possible and to use functions that use logic to find and replace records. For example:
```{r, c11, echo=T, results='hide'}

df[is.na("domSpp")]<-"white_oak"
```
Here R is looking for a value in the domSpp column that is NA (R's way of calling it a blank value). Since there's only 1, this works. If there were more than 1 NA, we'd need better logic to assign the proper value to each blank. We'll go into more detail about this when we're actually working with our forest data later.

And now, here's the tidyverse way to replace the blank in df's domSpp column with "white_oak":
```{r, c12}
df$domSpp <- df$domSpp %>% replace_na("white_oak")

```
We'll talk more about pipes (%>%) later, but for now, just know that it's taking the data on the left of the %>% and doing the thing on the right to it.

Finally, check whether plot is now reading in as numeric or still chr.
```{r, c13, echo=T}
str(df) #still a char
```
R still thinks plot is a char. The as.numeric function fixes the issue. If you didn't fix the last record, the plot column will still convert to numeric, but the last value will be set to NA, and you'll get a warning message: NAs introduced by coercion.
```{r, c14, echo=T}
df$plot <- as.numeric(df$plot)
str(df)
```

<p class='ques'>Question 2: How would you make numtrees an integer? </p>
</details>
</br>